### 2024-03-13 doc/15yrsChange.Rmd
text results, text changes, fig 6 changes

**Title:** Nuisance species compromise the carbon sequestration potential in an Eastern US temperate deciduous forest

@@

The temperate forest biome plays a critical role in the global carbon cycle, accounting for almost half of the net global forest carbon (C) sink [@harris_global_2021], with deciduous forests representing a substantial portion of this, sequestering >300 Tg C yr^-1^ [@pugh_role_2019]. 
Although currently a C sink, the future of the biome remains uncertain.
The dominant view imbued in Earth System Models is that, at least within the eastern US, the biome is likely to remain a C sink for the remainder of the 21st century [@finzi_carbon_2020; @wu_uncertainty_2023], albeit with declining CO~2~ sequestration capacity [@ahlström_robustness_2012]. 
Yet, global C models predict a wide range of future trajectories of CO~2~ sequestration [@ahlström_robustness_2012;@arora_carbon_2020]<!--check refs-->, and current global C models do not represent some influential mechanisms [e.g., @fatichi_moving_2014; @clark_continentwide_2021].
One mechanism that is not represented in global C models is the impact of nuisance species -- i.e., endemic or non-endemic species that, as a result of human influence, have much greater abundance in an ecosystem than they did historically, resulting in undesirable ecological consequences [@refs]. 
Similarly, the impact of nuisance species on forest carbon budgets is not considered in machine learning/niche models that seek to project future forest distribution and carbon stocks [**?** @wu_uncertainty_2023], nor in carbon offset projects [**???** @refs].
This is problematic in that nuisance species -- including nonindigenous insect pests and pathogens, nonindigenous plants, and over-abundant herbivores -- are dramatically impacting carbon cycling in temperate deciduous forests around the world, as described below.
"Tree mortality is one of the largest uncertainities in projections of future forest dynamics" (Bugmann et al. 2019)

Nuisance pests and pathogens can have a protracted and enduring impact on forests, often modifying forest productivity, nutrient cycling, successional trajectories and geomorphic processes (Lovett et al. 2016, Fischer et al. 2013, Herms and McCullough 2014, Fei et al. 2014).  The presence of nonindigenous forest insects and diseases, which have been historically introduced to new ecosystems through intercontinental trade, will continue to proliferate with increased globalization and human movement (Fischer et al. 2013, Aukema et al. 2010, Brockerhoff et al. 2006). Many of the prominent pests and pathogens in temperate deciduous forests have devastating effects on canopy composition and structure.  Once abundant in temperate deciduous forests throughout North America, the American Chestnut (*Castanea dentata*) and the Ash genera (*Fraxinus spp.*) have been decimated by pests and pathogens, with stand-clearing disturbances occuring within a few years after initial infestation (Anderson 2020, Klooster et al. 2013, Need a source for Ash abundance and length of time between infestation and mortality for Chestnuts). These previously dominant overstorey species are now confined to the understory as non-reproductive seeds and saplings (Elliot and Swank 2008, Anagnostakis 1987, Need a source for Ash).  In other species, such as the Eastern Hemlock (*Tsuga canadensis*), the existence of pests and pathogens induces close to total population mortality, with considerable risk of regional extirpation (Ellison et al. 2018).  The loss of mature canopy species to nuisance species contributes to global change driven biomass losses, with biomass losses due to pest and pathogens likely comparable to those attributed to fire and forest removal (Fei et al. 2013)  Since many pests and pathogens do not currently occupy the full extent of their host species range, a significant portion of the remaining forest canopy is susceptible to future invasion from nuisance species, further increasing potentials for biomass loss and altered carbon cycle dynamics (Fei et al. 2013)

Non-endemic pests and pathogens have been important driver of mortality [@anderson-teixeira_longterm_2021].
These can reduce C and need to be considered in future climate change projections.

@fei_biomass_2019

Non-endemic pests and pathogens facilitate invasion of non-endemic plants [@guo_impacts_2023 and refs therein].

>>>>

**Title:** Nuisance species compromise carbon sequestration potential in an Eastern US temperate deciduous forest

@@

The temperate forest biome plays a critical role in the global carbon cycle, accounting for almost half of the net global forest carbon (C) sink [@harris_global_2021], with temperate deciduous forests alone sequestering >300 Tg C yr^-1^ [@pugh_role_2019]. 
Although currently a net C sink, the future of the temperate forest biome remains uncertain.
The dominant pattern projected by Earth System Models is a continued, albeit weakening, temperate forest C sink for the remainder of the 21st century [@wu_uncertainty_2023; @ahlström_robustness_2012]. 
Yet, global C models predict a wide range of future trajectories of CO~2~ sequestration [@ahlström_robustness_2012;@arora_carbon_2020]<!--check refs-->, and current global C models do not represent some influential mechanisms [e.g., @fatichi_moving_2014; @clark_continentwide_2021].
One mechanism that is not represented in global C models is the impact of nuisance species – i.e., indigenous or non-indigenous species that, as a result of human influence, have much greater abundance in an ecosystem than they did historically, resulting in undesirable ecological consequences [@refs]. 
Similarly, the impact of nuisance species on forest carbon budgets is not considered in machine learning or niche models that seek to project future forest distribution and carbon stocks, nor in carbon offset projects [@wu_uncertainty_2023].
This is problematic in that nuisance species – including non-indigenous insect pests and pathogens, non-indigenous plants, and over-abundant herbivores – are dramatically impacting carbon cycling in temperate deciduous forests around the world.

Non-indigenous pests and pathogens can have a protracted impact on forests, often modifying forest productivity, nutrient cycling, successional trajectories, and geomorphic processes (Lovett et al. 2016, Fischer et al. 2013, Herms and McCullough 2014, Fei et al. 2014). Historically introduced to new ecosystems through intercontinental trade, non-indigenous forest insects and diseases will continue to proliferate with increased globalization and human movement (Fischer et al. 2013, Aukema et al. 2010, Brockerhoff et al. 2006). Many of the prominent pests and pathogens in temperate deciduous forests have dramatic effects on canopy composition and structure, decimating once-abundant canopy species (Aukema et. al 2010,Anderson-Teixeria 2021, Klooster et al. 2013, Ellison et al. 2018). Since many pests and pathogens do not currently occupy the full extent of their host species range, a significant portion of the remaining forest canopy is susceptible to future invasion from nuisance species, further increasing potentials for biomass loss and altered carbon cycle dynamics (Fei et al. 2013).
